Estimating SARS-CoV-2 transmission in England from randomly sampled prevalence surveys

### 2023-12-04 ATACCC-appendix-original-analysis.tex
Concrete ATACCC analyses

\todo[inline]{Check for "REF" in this appendix}

@@

Inference was performed by Hamiltonian Monte Carlo (HMC) implemented in Stan 2.26.1~[REF].
To aid with mixing, the hierarchical parameters were implemented using the non-centred parameterisation~[REF].
In addition, $t_{i,0}$ was reparameterised in terms of the absolute time of peak.
Convergence was assessed by checking R-hat < 1.1, and visual inspection of traceplots, and effective sample size.

>>>>

@@

Inference was performed by Hamiltonian Monte Carlo (HMC) implemented in Stan 2.26.1~\autocite{Stan-2-32-2}.
To aid with mixing, the hierarchical parameters were implemented using the non-centred parameterisation~\autocites{papaspiliopoulosGeneral,stanReparameterization}.
In addition, $t_{i,0}$ was reparameterised in terms of the absolute time of peak.
Convergence was assessed by checking R-hat < 1.1, and visual inspection of traceplots, and effective sample size.

### 2023-12-04 ATACCC.tex
Concrete ATACCC analyses

Several features of the \textcite{hakkiOnset} model can be motivated by the desire to apply Stan, a modern and convenient MCMC implementation of the NUTS algorithm (see\todo{ref wherever I end up describing NUTS}).
The first of these is using a continuous viral load trajectory rather than the more common piecewise linear model.
As NUTS uses gradient information, it is more efficient when the model is differentiable.
This leads to the following model for the viral load at time $t$, $v_i(t)$, in individual $i$ is shown in \cref{ATACCC:fig:viral-load-model} and given by:
\begin{align}
v_i(\tau) = v_{\max,i} \frac{a_i+b_i}{b_ie^{-a_i(\tau-\tau_{\max,i})} + a_ie^{b_i(\tau-\tau_{max,i})}}. \label{ATACCC:eq:viral-load}
\end{align}
Time, $\tau$, is continuous and defined uniquely for each individual such that $\tau = 0$ at the time that individual $i$ was first tested.
$v_{\max,i}$ is the maximum viral load, which occurs at time $\tau_{\max,i}$.
As $\lvert \tau - \tau_{\max} \rvert$ becomes large, the viral load trajectory tends towards exponential growth (before the peak) or decline (after the peak) in viral load at rates $a_i$ and $b_i$ respectively.
% Before the peak, this is because $b(\tau-\tau_{\max,i})$ is small and hence $a\exp(b(\tau-\tau_{\max,i}))$ is negligible.
% After the peak, this is because $b\exp(-a(\tau-\tau_{\max,i}))$ is negligible.
This happens quickly for realistic values of the parameters.
Therefore, the log viral load is approximately piecewise linear (see \cref{ATACCC:fig:viral-load-model} for a representative viral load trajectory).
$\vec{\phi}_i = (v_{\max,i}, a_i, b_i)^T$ are the random effects for individual $i$.
$\tau_{\max,i}$ is inferred independently for each individual because the time from first test to peak is not a biological concept shared between individuals.
\begin{figure}
  \centering \includegraphics{ATACCC/typical_trajectory}
  \caption[Example viral load model trajectory.]{Example viral load trajectory produced by \textcite{hakkiOnset}'s viral load model, using realistic parameter values. The increase in log viral load is approximately linear with a gradient of $a_i$, here 5. The decrease is also linear, with gradient $-b_i$, here -1.4. The peak is at time $\tau_{\max,i}$, here 0, and log viral load $\log v_{\max,i}$, here = 15. The smoothing (lack of linearity) near the peak is visible. \label{ATACCC:fig:viral-load-model}}
  \todo[inline]{Add a few lines to example plot showing the effect of different parameters, move parameters to key}
\end{figure}

PCR tests are noisy observations of the viral load.
A positive PCR test has a Ct value.
Ct values are linearly related to viral load (see\todo{ref intro section}).
A negative PCR test signals either the viral load is too low to be detected, \ie below the limit of detection, or the test is a false negative.
Positive PCR tests, especially those with low viral loads, could be false positives, for example caused by contamination.

@@

\todo[inline]{insert a better explanation of $F_D$'s posterior}
For later use, a discretised version is preferred (as described in \cref{E-inc-prev}).
To calculate the discrete version of this distribution, $F_D$, a Monte Carlo integration is used with the following relation.
\begin{align}
  F_D(d)
  &= \prob(D_i \leq d + 0.5) \\
  &= \E_{D_i}(\I(D_i \leq d + 0.5)) 
\end{align}
where $\I$ is the indicator function.
Here, $D_i$ is drawn conditional on the population-level parameters $\vec{\mu}_\phi$ and $\matr{\Sigma}_\phi$.
Equivalent forms of $F_D$, such as the survival and hazard functions, can be calculated using the relations in\todo{ref section defining survival functions}.
I found 100,000 draws produced stable estimates of the tail of $F_D$.

@@

\ifSubfilesClassLoaded{
  \listoftodos
  \appendix
  \subfile{distributions}
  \subfile{ATACCC-appendix-original-analysis}
  \listoftodos
}{}

>>>>

Several features of the \textcite{hakkiOnset} model can be motivated by the desire to apply Stan, a modern and convenient MCMC implementation of the NUTS algorithm (see \cref{E-backcalc:sec:MCMC}).
The first of these is using a continuous viral load trajectory rather than the more common piecewise linear model.
As NUTS uses gradient information, it is more efficient when the model is differentiable.
This leads to the following model for the viral load at time $t$, $v_i(t)$, in individual $i$ is shown in \cref{ATACCC:fig:viral-load-model} and given by:
\begin{align}
v_i(\tau) = v_{\max,i} \frac{a_i+b_i}{b_ie^{-a_i(\tau-\tau_{\max,i})} + a_ie^{b_i(\tau-\tau_{max,i})}}. \label{ATACCC:eq:viral-load}
\end{align}
Time, $\tau$, is continuous and defined uniquely for each individual such that $\tau = 0$ at the time that individual $i$ was first tested.
$v_{\max,i}$ is the maximum viral load, which occurs at time $\tau_{\max,i}$.
As $\lvert \tau - \tau_{\max} \rvert$ becomes large, the viral load trajectory tends towards exponential growth (before the peak) or decline (after the peak) in viral load at rates $a_i$ and $b_i$ respectively.
% Before the peak, this is because $b(\tau-\tau_{\max,i})$ is small and hence $a\exp(b(\tau-\tau_{\max,i}))$ is negligible.
% After the peak, this is because $b\exp(-a(\tau-\tau_{\max,i}))$ is negligible.
This happens quickly for realistic values of the parameters.
Therefore, the log viral load is approximately piecewise linear (see \cref{ATACCC:fig:viral-load-model} for a representative viral load trajectory).
$\vec{\phi}_i = (v_{\max,i}, a_i, b_i)^T$ are the random effects for individual $i$.
$\tau_{\max,i}$ is inferred independently for each individual because the time from first test to peak is not a biological concept shared between individuals.
\begin{figure}
  \centering \includegraphics{ATACCC/typical_trajectory}
  \caption[Example viral load model trajectory.]{Example viral load trajectory produced by \textcite{hakkiOnset}'s viral load model, using realistic parameter values. The increase in log viral load is approximately linear with a gradient of $a_i$, here 5. The decrease is also linear, with gradient $-b_i$, here -1.4. The peak is at time $\tau_{\max,i}$, here 0, and log viral load $\log v_{\max,i}$, here = 15. The smoothing (lack of linearity) near the peak is visible. \label{ATACCC:fig:viral-load-model}}
  \todo[inline]{Add a few lines to example plot showing the effect of different parameters, move parameters to key}
\end{figure}

PCR tests are noisy observations of the viral load.
A positive PCR test has a Ct value.
Ct values are linearly related to viral load (see\todo{make sure relationship between Ct values and viral load is introduced previously}).
A negative PCR test signals either the viral load is too low to be detected, \ie below the limit of detection, or the test is a false negative.
Positive PCR tests, especially those with low viral loads, could be false positives, for example caused by contamination.

@@

\todo[inline]{insert a better explanation of $F_D$'s posterior}
For later use, a discretised version is preferred (as described in \cref{E-inc-prev}).
To calculate the discrete version of this distribution, $F_D$, a Monte Carlo integration is used with the following relation.
\begin{align}
  F_D(d)
  &= \prob(D_i \leq d + 0.5) \\
  &= \E_{D_i}(\I(D_i \leq d + 0.5)) 
\end{align}
where $\I$ is the indicator function.
Here, $D_i$ is drawn conditional on the population-level parameters $\vec{\mu}_\phi$ and $\matr{\Sigma}_\phi$.
Equivalent forms of $F_D$, such as the survival and hazard functions, can be calculated using the relations in \cref{E-backcalc:sec:definitions}.
I found 100,000 draws produced stable estimates of the tail of $F_D$.

@@

\ifSubfilesClassLoaded{
  \appendix
  \subfile{distributions}
  \subfile{ATACCC-appendix-original-analysis}
  \listoftodos
}{}
